In this section, I describe the major components of the MIPS
processor. I look at the components before wiring them together into a
processor, as it keeps each step more isolated and simple. The source code for
each of the components is available at~\cite{ref:github} in
\texttt{sme/src/Examples/SingleCycleMIPS/}.

For each component, I will go through the theory of the component, then I
will look at translating theory into an SME implementation. To verify the
implementation, I construct a tester process, just as with the basic logic
circuits.  The order in which the components are mentioned is derived from
chapter 4.3 in the architecture book~\cite{ref:ark}, and should be the order of
implementation.

By the end of this section, I should have all of the required resources for
building the single cycle MIPS processor using SME.

\subsection{Instruction Memory}
The first component that I need is a memory unit to hold the instructions of a
program, and to supply instructions at a given address. I will also need a
register holding the current address in the program called the Program Counter
(PC). The memory unit has one input: the address from the PC, and one output:
the read instruction. The PC has one input: the input address. The Instruction
Memory, the PC and their connections can be seen in Figure~\ref{fig:inst}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (inmem) at (0,0) {Instruction Memory};
        \node[block] (pc) at (-2, 2) {PC};
        \node[empty] (prog) at (1.5,2) {Program Counter};
        \node[empty] (inst) at (4,0) {Instruction};
        \node[empty] (addr) at (-4,2) {Address};

        \path[draw, ->] (pc) -| (inmem);
        \path[draw, ->] (inmem) -- (inst);
        \path[draw, ->] (addr) -- (pc);
    \end{tikzpicture}
    \caption{The Instruction Memory.}
    \label{fig:inst}
\end{figure}

\subsubsection*{Implementation}
Both the PC and the Instruction Memory should be SME processes. I could have
combined the two components into a single SME process, but as mentoined before,
I am trying to be closer to the CSP model. The input bus for the PC, the Program
Counter bus and the instruction bus, should all contain a \texttt{uint} value.

The PC should have a single \texttt{uint} value holding the current address.
It should only output its stored value, when the clock ticks. This can be
expressed in SME, by giving a process the \texttt{ClockedProcess} attribute.
This ensures that the process is no longer combinatorial, but only performs
its operation when the clock ticks.

The Instruction Memory should have an \texttt{byte} array, as this will make
indexing into the array simpler. Usually, the Instruction Memory and the Memory
Unit share the same address space. However, for simplicity, I give each
component its own memory. On each tick, the Instruction Memory should take the
input address, read the \texttt{byte} at the given index and the following 3
\texttt{byte}s, and finally pack all 4 \texttt{byte}s together into an
\texttt{uint}, and place it on the instruction bus.

I could have had the memory as an \texttt{uint} array, but then I would have
to divide the address by 4, since the memory in the MIPS processor are byte
addressed. I could modify the processor to deal with addresses in an indexing
manner, however this would make programs harder to translate, as the
instructions in a MIPS binary can have absolute addresses.

\subsection{Register file}
The MIPS processor has 32 general-purpose 32-bit registers, which are stored in
a structure called the Register File. Each of the registers can be read from or
written to, except for register zero, which is immutable and always 0. It is
the first level in a memory hierarchy, and is thus the fastest memory
available. The registers are divided into groups based on their usage, but
this does not matter from a hardware perspective.

The Register File has 5 inputs: Read Register A, Read Register B, Write Enabled
(\texttt{RegWrite}), Write Address and Write Data. The Register File has two
outputs: Output A and Output B.  There are two stages of the Register File:
reading and writing. In the reading stage, it takes the value in the register
at the address of Read Register A, and outputs it on Output A, and vice versa
for Read Register B and Output B. In the writing stage, if the Write Enabled
flag is set, it takes the value from the Write Data bus, and stores it in the
register with the address given in the Write Address bus.

I need to be careful of the order in which I read and write from the Register
File. I need to make sure that when an instruction reads from the Register
File, it always gets the latest data (i.e. if an instruction reads from the
same register as a previous instruction writes to, it should read the value
written by the previous instruction). This is easy to fix in the single cycle
processor, as the register file should just write before reading. It should not be
in the reverse order, as the Register File might then output old values. The
Register File and its connections can be seen in Figure~\ref{fig:register}.

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=2cm]
        \node[empty] (inputa) {Read Register A };
        \node[empty, below of=inputa] (inputb) {Read Register B};

        \node[empty, right of=inputa] (spacing) at ($(inputa)!0.5!(inputb)$) {};
        \node[block, right of=spacing] (register) {Register};
        \node[empty, below of=register] (writedata) {Write Data};
        \node[empty, left of=writedata] (write) {Write Address};
        \node[empty, right of=writedata] (writeenabled) {Write Enabled};

        \node[empty, right of=inputa] (space) {};
        \node[empty, right of=space] (spacee) {};
        \node[empty, right of=spacee] (spaceee) {};
        \node[empty, right of=spaceee] (outputa) {Output A};
        \node[empty, below of=outputa] (outputb) {Output B};
        \node[empty, left of=outputb] (bspace) {};

        \path[draw, -] (inputa) -| (spacing.north);
        \path[draw, ->] (spacing.north) |- (register.155);
        \path[draw, -] (inputb) -| (spacing.south);
        \path[draw, ->] (spacing.south) |- (register.205);

        \path[draw, color=blue, -] (write) |- (writedata.135);
        \path[draw, color=blue, ->] (writedata.135) -- (register.225);
        \path[draw, color=blue, ->] (writedata) -- (register);
        \path[draw, color=blue, -] (writeenabled) |- (writedata.45);
        \path[draw, color=blue, ->] (writedata.45) -- (register.315);

        \path[draw, -] (register.335) -| (bspace.north);
        \path[draw, ->] (bspace.north) |- (outputb);
        \path[draw, -] (register.25) -| (spaceee.south);
        \path[draw, ->] (spaceee.south) |- (outputa);
    \end{tikzpicture}
    \caption{The register file.}
    \label{fig:register}
\end{figure}

\subsubsection*{Implementation}
The Register File should be an SME process. The collection of registers should
be a \texttt{uint} array of length 32. The address busses should all contain
a \texttt{byte} value, as the number of addressable registers never exceeds
$2^8=256$. The output busses and the Write Data bus should all contain a
\texttt{uint} value. Finally, the \texttt{RegWrite} bus should contain a
\texttt{bool}.

On each clock tick, the Register File should check if the
\texttt{RegWrite} flag is set, in which case it should take the value from the
Write Data bus, and store in the register at the address from the Write Address
bus. Then it should take the value in the register at the address from the Read
Register A bus, and output onto the Output A bus, and similarly for Read
Register B and Output B. It should be noted that if the write address is 0, then
the write should be ignored, as register zero is immutable. The simplified code
for the Register File SME process can be seen in Listing~\ref{lis:reg}.

\begin{figure}
\begin{lstlisting}[caption=The SME process for the register file., label=lis:reg, captionpos=b]
public class Register : SimpleProcess
{
    ...

    uint[] data = new uint[32];

    protected override void OnTick()
    {
        if (regWrite.flg && writeAddr.val > 0)
            data[writeAddr.val] = writeData.data;
        outputA.data = data[readA.addr];
        outputB.data = data[readB.addr];
    }
}
\end{lstlisting}
\end{figure}

\subsection{ALU}
\label{sec:alu}
The ALU (Arithmetic Logic Unit) is the part of the processor, which makes the
actual computation. It takes three inputs: Input A, Input B and an ALU Opcode
indicating which computation to perform. It has two outputs: The result of the
computation, and a zero flag indicating whether or not the result of the
computation was 0. The ALU and its connections can be seen in
Figure~\ref{fig:alu}.

\begin{figure}
    \centering
    \begin{tikzpicture} [node distance=1.5cm]
        \node[empty] (ina) {Input A};
        \node[empty, below of=ina] (inb) {Input B};
        \node[empty, right of=inputa] (spacing) at ($(ina)!0.5!(inb)$) {};
        \node[block, right of=spacing] (alu) {ALU};
        \node[empty, right of=ina] (align0) {};
        \node[empty, right of=align0] (align1) {};
        \node[empty, right of=align1] (align2) {};
        \node[empty, right of=align2] (result) {Zero};
        \node[empty, below of=result] (zero) {Result};
        \node[empty, left of=zero] (align3) {};
        \node[empty, below of=alu] (aluop) {ALUOp};

        \path[draw, -] (ina.east) -| (spacing.155);
        \path[draw, ->] (spacing.155) -- (alu.155);
        \path[draw, -] (inb.east) -| (spacing.205);
        \path[draw, ->] (spacing.205) -- (alu.205);
        \path[draw, color=blue, ->] (aluop) -- (alu);
        \path[draw, color=blue, -] (alu.25) -| (align2.east);
        \path[draw, color=blue, ->] (align2.east) -- (result);
        \path[draw, -] (alu.335) -| (align3.east);
        \path[draw, ->] (align3.east) -- (zero);
    \end{tikzpicture}
    \caption{The ALU.}
    \label{fig:alu}
\end{figure}


The ALU starts by looking at the value in the ALU Opcode, as this determines
which operation to perform. Then it reads the values from Input A and Input B,
and performs the operation specified by the ALU Opcode. Finally, it outputs the
result, and sets the zero flag to whether the result was 0.

\subsubsection*{Implementation \& Testing}
To implement the ALU, I start by making an \texttt{enum}, so that the code
becomes more human readable. Each entry in the \texttt{enum} corresponds to a
computation that the ALU should perform. I could have followed the ALU Opcode
given in the architecture book~\cite{ref:ark}, however it states that its
encoding of the ALU Opcode is generated using a CAD tool, and as such I will
let the VHDL transpiler generate it too.

I start by implementing the same instructions as the architecture
book~\cite{ref:ark} proposes in chapter 4.1: \texttt{add}, \texttt{sub},
\texttt{and}, \texttt{or}, \texttt{slt}, \texttt{sw}, \texttt{lw} and
\texttt{beq}. To perform these instructions, the ALU should be able to perform
addition, subtraction, bitwise \texttt{AND}, bitwise \texttt{OR} and the
comparison less than.

The two input busses and the Result bus should all contain a \texttt{uint}
value, the ALU Opcode bus should contain a \texttt{byte} value and the Zero bus
should contain a \texttt{bool} value.

Constructing the ALU process in SME is straightforward, it reads the ALU Opcode
from the ALU Opcode bus, and then it performs a \texttt{switch} on the ALU Opcode.
For the instructions that it accepts, it takes the input from the two input
busses, do the computation, and output the result on the Result bus. SME will
handle the the actual computation, and as such I do not need to implement the
ALU circuitry, but rather do the right computation in C\#. Note: it is
important to cast the \texttt{uint} input to \texttt{int}, if the operation is
signed.

Finally, the ALU should set the flag on the Zero bus, depending on whether or
not the result of the computation was 0. How the ALU opcode is encoded is
described in Section~\ref{sec:alu-control}.

\subsection{Sign Extend}
The Sign Extend is used for extracting the 16-bit values from the instruction,
called the Immediate. It has one input: the lower 16 bits from the
instruction, and one output: the 32-bit sign extended value. It takes its
input, which is 16-bit, and converts it into a 32-bit value, extending the sign
if present. The Sign Extend and its connections can be seen in
Figure~\ref{fig:sign}.

\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (sign) at (0,0) {Sign Extend};
        \node[empty] (imm) at (-3,0) {Immediate};
        \node[empty] (immout) at (4,0) {Sign extended immediate};

        \path[draw, ->] (imm) -- (sign);
        \path[draw, ->] (sign) -- (immout);
    \end{tikzpicture}
    \caption{The Sign Extend.}
    \label{fig:sign}
\end{figure}

\subsubsection*{Implementation}
To implement the Sign Extend, I construct an SME process. The input bus should
contain a \texttt{short} value, and the output bus should contain a
\texttt{uint} value. On each clock tick, the SME process takes the 16-bit
immediate, and outputs it on its 32-bit output bus. I could manually do the
sign extending, but I just let C\# handle the sign extension.

\subsection{Memory unit}
The Memory Unit is the main memory of the processor. It is the second step of
the memory hierarchy, and is thus slower than the Register File, but can
contain a lot more data. The processor can either read or write to the Memory
Unit in a single cycle. The addresses for the memory are byte addresses and
word aligned (i.e. in the 32-bit processor, the word size is 32 bit) and
therefore the address should be divisable by 4.

The Memory Unit has four inputs: Address, Data, \texttt{MemRead} and
\texttt{MemWrite}. It has a single output: Read Data. In one clock cycle, the
Memory Unit either reads or writes. The Memory Unit and its connections can be
seen in Figure~\ref{fig:mem}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (mem) at (0,0) {Memory};
        \node[empty] (addr) at (-3,0.4) {Address};
        \node[empty] (data) at (-3,-0.4) {Data};
        \node[empty] (memread) at (-1.5, 1) {\texttt{MemRead}};
        \node[empty] (memwrite) at (1.5, 1) {\texttt{MemWrite}};
        \node[empty] (readdata) at (3, 0) {Read Data};

        \path[draw, ->] (addr) -- (mem.154);
        \path[draw, ->] (data) -- (mem.206);
        \path[draw, color=blue, ->] (memread) -| (mem.110);
        \path[draw, color=blue, ->] (memwrite) -| (mem.70);
        \path[draw, ->] (mem) -- (readdata);
    \end{tikzpicture}
    \caption{The Memory Unit.}
    \label{fig:mem}
\end{figure}

\subsubsection*{Implementation}
To implement the Memory Unit, I construct an SME process. The Address, Data
and Output busses should all contain a \texttt{uint} value. The two control
busses \texttt{MemRead} and \texttt{MemWrite} should both contain a
\texttt{bool} value.

As with the Instruction Memory, it is going to need a chunk of memory. Like
the Instruction Memory, the memory chunk should be a \texttt{byte} array, and
should read and write in the same manner, i.e. either packing or unpacking 4
\texttt{byte}s from and to memory.

On each clock tick, the process should check if the \texttt{MemRead} flag is
set, in which case it should read the value on the Address bus, and output the
value stored in memory at the read address. If that was not the case, it should
check if the \texttt{MemWrite} flag is set, in which case it should read the
value at the Address bus and the Data bus, store the read data in memory at the
read address.

\subsection{Splitter}
The architecture book~\cite{ref:ark} takes the instruction read from the
Instruction Memory, and splits it out to the different components. One way of
handling this could be to send the instruction bus to all the components that
needs something from the instruction. However, then the resulting VHDL might
send all 32 bits to multiple components, when 5 bits might have been
sufficient. As such, I construct my own component for splitting up the
instruction.

The splitter is a very simple component: It takes the instruction, which has
been fetched from memory, and divides it into chunks for the different parts of
the decoding. The instruction is partitioned as followed (the bits are
inclusive):
\begin{itemize}
    \item Opcode - bits 26-31
    \item Read Address A - bits 21-25
    \item Read Address B - bits 16-20
    \item Write Address - bits 11-15
    \item Immediate - bits 0-15
    \item Funct - bits 0-5
\end{itemize}
The Splitter and its connections can be seen in Figure~\ref{fig:split}
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (split) at (0,0) {Splitter};
        \node[empty] (inst) at (-3,0) {Instruction};
        \node[empty] (op) at (4,1.25) {Opcode};
        \node[empty] (ra) at (4,0.75) {Read Address A};
        \node[empty] (rb) at (4,0.25) {Read Address B};
        \node[empty] (wa) at (4,-0.25) {Write Address};
        \node[empty] (imm) at (4,-0.75) {Immediate};
        \node[empty] (fun) at (4,-1.25) {Funct};

        \path[draw, ->] (inst) -- (split);
        \path[draw, ->] (split.25) -| (1.5, 1) |- (op);
        \path[draw, ->] (split.15) -| (1.75,0.5) |- (ra);
        \path[draw, ->] (split.5) -| (2,0.125) |- (rb);
        \path[draw, ->] (split.355) -| (2,-0.125) |- (wa);
        \path[draw, ->] (split.345) -| (1.75,-0.5) |- (imm);
        \path[draw, ->] (split.335) -| (1.5, -1) |- (fun);
    \end{tikzpicture}
    \caption{The Splitter.}
    \label{fig:split}
\end{figure}


\subsubsection*{Implementation}
The implementation is straightforward: I construct an SME process, which takes
the instruction coming from the Instruction Memory, and extract the bits at the
indices. It would be nice to be able to split a signal, in the same manner as
VHDL does it (i.e. by having a slice index), however SME does not support this.
Instead, I use use C\# bit hacking, by shifting and masking the number, as this
has the semantics as splitting a signal. The Input bus should contain an
\texttt{uint} value. The Immediate bus should contain a \texttt{short} value.
The rest of the busses should contain a \texttt{byte} value. Finally, the
process should output the extracted values on the respective output busses.

\subsection{Jump Unit}
The architecture book~\cite{ref:ark} describes how to implement \texttt{beq},
by the use of multiple components. However, I combine these components into a
single unit, the Jump Unit, to simplify the processor. The unit should still be
composed of the subcomponents as suggested by the architecture
book~\cite{ref:ark}.

The Jump Unit is the one controlling which instruction to load next. It takes
four inputs: sign extend, Zero, \texttt{beq} and the PC. It produces one
output: the new PC.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (and) at (0,0) {\texttt{AND}};
        \node[block] (inc) at (-2,1) {+4};
        \node[block] (add) at (2,1) {+};
        \node[empty] (pc) at (-4,1) {PC};
        \node[empty] (beq) at (-0.5,-1.5) {\texttt{beq}};
        \node[empty] (zero) at (0.5,-1.5) {Zero};
        \node[empty] (imm) at (4,1) {Immediate};
        \node[empty] (out) at (-4,2) {New PC};
        \node[mux] (mux) at (0,2) {|};

        \path[draw, ->, color=blue] (beq.65) -- (and.245);
        \path[draw, ->, color=blue] (zero.115) -- (and.295);
        \path[draw, ->, color=blue] (and) -- (mux);
        \path[draw, ->] (pc) -- (inc);
        \path[draw, ->] (inc) -- (add);
        \path[draw, ->] (imm) -- (add);
        \path[draw, ->] (add.north) |- (mux.45);
        \path[draw, ->] (1, 1) |- (mux.315);
        \path[draw, ->] (mux) -- (out);

        \draw (-2.8,-0.8) rectangle (2.8,2.8);
    \end{tikzpicture}
    \caption{The Jump Unit.}
    \label{fig:jump}
\end{figure}

For the simple single cycle MIPS processor, it should only have support for
normal program traversal (i.e. execute the instructions in order) and the
\texttt{beq} instruction. I will be adding support for more branch and jump
instructions later in Section~\ref{sec:extending}.

For the simple traversal, the Jump Unit takes the previous PC, and increments
it by 4. For the \texttt{beq} instruction, it takes the value from the Sign
Extend, shifts it left by 2, and add that value to the incremented PC. Finally,
it chooses between the incremented PC and the added sign extend, based on the
\texttt{beq} and Zero flags. The Jump Unit, its connections and its internals
can be seen in Figure~\ref{fig:jump}.

\subsubsection*{Implementation}
I have the all of the logic described in the background. To implement the Jump
Unit, I implement all of the logic components as SME processes. I am going
to need 4 components: an +4 incrementer, an adder, a \texttt{AND} gate and a
multiplexor. I could have made all the logic in a single process. However this
makes it easier to extend, increases concurrency and is closer to the CSP
model, as each subcomponent is very simple.

The Incrementer takes the PC bus as input, and produces a single output: the
incremented bus. The two busses should both contain a \texttt{uint} value. On
each clock tick, it should take the value from the PC bus, add 4 to it and put
it on the incremented bus.

The Adder takes the inc bus and the immediate bus from the Sign Extend as
input. It has a single output bus: the added bus. All of the busses should
contain a \texttt{uint} value. On each clock tick, the Adder should add its
two inputs together, and put the result on the add bus.

The \texttt{AND} gate is like in the Logic Gates example, but with the
\texttt{beq} and Zero as inputs, and its output should be on a new bus: anded.
All of its busses should contain a \texttt{bool} value.

Finally, I have the multiplexor, which takes the incremented, added and anded
busses as input, and produces a single output: the new PC. The new PC bus
should contain a \texttt{uint} value. On each clock tick, if the flag from the
anded bus is set to \texttt{1}, then it should put the value from the added bus
onto the new PC bus. Otherwise it should output the value from the inced bus.

\subsection{ALU control}\label{sec:alu-control}
The ALU control is used for generating the ALU Operation control code, which
the ALU uses for selecting which operation to perform. It takes two inputs: the
\texttt{ALUOp} code from the control unit, and the \texttt{funct} code from the
instruction. It produces a single output: The ALU Opcode. The ALU Control  and
its connections can be seen in Figure~\ref{fig:alu-cont}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (alucont) at (0,0) {ALU Control};
        \node[empty] (aluop) at (-3, -1) {ALUOp};
        \node[empty] (funct) at (-3, -1.5) {Funct};
        \node[empty] (aluope) at (0, 2) {ALU Operation};

        \path[draw, ->, color=blue] (aluop) -| (alucont.250);
        \path[draw, ->] (funct) -| (alucont.290);
        \path[draw, ->, color=blue] (alucont) -- (aluope);
    \end{tikzpicture}
    \caption{The ALU Control.}
    \label{fig:alu-cont}
\end{figure}

The MIPS processor accepts three basic instruction formats~\cite{ref:ark-book}:
The R format, the I format and the J format. The R format instructions are
decoded in the ALU Control. The I and J formats are decoded in the Control
Unit. The R format is divided into six parts:
\begin{itemize}
    \item \makebox[1.3cm]{\texttt{opcode}\hfill}
        \makebox[1.3cm]{(31-26)\hfill}: Opcode of the instruction
    \item \makebox[1.3cm]{\texttt{rs}\hfill}
        \makebox[1.3cm]{(25-21)\hfill}: Source register 1
    \item \makebox[1.3cm]{\texttt{rt}\hfill}
        \makebox[1.3cm]{(20-16)\hfill}: Source register 2
    \item \makebox[1.3cm]{\texttt{rd}\hfill}
        \makebox[1.3cm]{(15-11)\hfill}: Destination register
    \item \makebox[1.3cm]{\texttt{shamt}\hfill}
        \makebox[1.3cm]{(10-6)\hfill}:
        Shift amount (only used by shift instructions)
    \item \makebox[1.3cm]{\texttt{funct}\hfill}
        \makebox[1.3cm]{(5-0)\hfill}:   Function opcode
\end{itemize}

If the \texttt{ALUOp} indicates that the instruction is an R format
instruction, it uses the \texttt{funct} code in the instruction for selecting
the operation. Otherwise it bases its output on the \texttt{ALUOp} code. I will
return to this component, when I need to extend the instruction set in
Section~\ref{sec:extending}.

\subsubsection*{Implementation}
The architecture book~\cite{ref:ark-book} describes the logic to implement the
ALU Control. However, like the ALU, it has been generated, and is difficult to
extend. As such, I construct my own ALU Control, and let the VHDL transpiler
handle the logic generation.

To make the source code more human readable, I have two additional
\texttt{enum}s: one for the \texttt{ALUOp} code and one for the \texttt{funct}.
Then I construct an SME process, which has the connections as specified in the
background section. All of its busses should contain \texttt{byte} values.

On each clock tick, the ALU Control checks if the \texttt{ALUOp} code indicate
R format instruction or not. If the instruction is an R format, the process
should \texttt{switch} on the \texttt{funct} code. If not, it should
\texttt{switch} on the \texttt{ALUOp} code. In all cases in both
\texttt{switch}s, the ALU Control should output the ALU Operation corresponding
to the computation that the instruction expects.

\subsection{Control Unit} \label{sec:control-unit}
The control unit is part of the decoding step. It takes the opcode of the
instruction, and based on the opcode, it sets control flags used throughout the
processor. All of the control flags mentioned throughout the other components
are set by the Control Unit. It sets the following control flags:\\\\
\begin{tabular}{lp{9cm}}
    \bf{RegDst} & Controls which part of the instruction that indicates the
        Read Address B to the Register File. \\

    \bf{Branch} & Controls whether or not the instruction is a branch
        instruction. \\

    \bf{MemRead} & Controls whether or not there should be read from memory. \\

    \bf{MemtoReg} & Controls whether or not the value from memory should be
        stored in the register file. \\

    \bf{ALUOp} & Opcode indicating which operation should be performed in
        the ALU. It is sent to the ALU Control for further processing. \\

    \bf{MemWrite} & Controls whether or not data should be written to memory. \\

    \bf{ALUSrc} & Controls whether the B input for the ALU should be the
        value read from the register file, or if it should be the value
        extracted from the instruction. \\

    \bf{RegWrite} & Controls whether or not data should be written to the
        register file. \\
\end{tabular}\\

As mentioned in Section~\ref{sec:alu-control}, the Control Unit decodes I and J
format instructions. The J format instructions are described in
\textit{Jump instruction} in Section~\ref{sec:extending}. The I instructions
are the immediate instructions, i.e. instructions where one of the source
values are contained within the instruction~\cite{ref:ark-book}. An I format
instruction is divided into four parts:
\begin{itemize}
    \item \makebox[1.8cm]{\texttt{opcode}\hfill}
        \makebox[1.3cm]{(31-26)\hfill} Opcode of the instruction
    \item \makebox[1.8cm]{\texttt{rs}\hfill}
        \makebox[1.3cm]{(25-21)\hfill} Source register
    \item \makebox[1.8cm]{\texttt{rd}\hfill}
        \makebox[1.3cm]{(20-16)\hfill} Destination register
    \item \makebox[1.8cm]{\texttt{immediate}\hfill}
        \makebox[1.3cm]{(15-0)\hfill} Immediate source
\end{itemize}

Each of the control flags goes to their respective part of the processor.
I will return to this component, when I have to extend it to handle more
instructions in Section~\ref{sec:extending}. The control unit and its
connections can be seen in Figure~\ref{fig:cont-unit}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[control] (cont)   at (0,  0)   {Control Unit};
        \node[empty] (opcode)   at (-3, 0)   {Opcode};
        \node[empty] (branch)   at (4,  1.75) {\texttt{Branch}};
        \node[empty] (regwrite) at (4,  1.25) {\texttt{RegWrite}};
        \node[empty] (memtoreg) at (4,  0.75) {\texttt{MemToReg}};
        \node[empty] (memwrite) at (4,  0.25) {\texttt{MemWrite}};
        \node[empty] (memread)  at (4, -0.25) {\texttt{MemRead}};
        \node[empty] (alusrc)   at (4, -0.75) {\texttt{ALUSrc}};
        \node[empty] (aluop)    at (4, -1.25) {\texttt{ALUOp}};
        \node[empty] (regdst)   at (4, -1.75) {\texttt{RegDst}};

        \path[draw, ->] (opcode) -- (cont);

        \path[draw, ->, color=blue] (cont.35)  -| (2,1.5) |- (branch);
        \path[draw, ->, color=blue] (cont.25)  -| (2.25,1) |- (regwrite);
        \path[draw, ->, color=blue] (cont.15)  -| (2.5,0.5) |- (memtoreg);
        \path[draw, ->, color=blue] (cont.5)   -| (2.75,0.125) |- (memwrite);
        \path[draw, ->, color=blue] (cont.355) -| (2.75,-0.125) |- (memread);
        \path[draw, ->, color=blue] (cont.345) -| (2.5,-0.5) |- (alusrc);
        \path[draw, ->, color=blue] (cont.335) -| (2.25,-1) |- (aluop);
        \path[draw, ->, color=blue] (cont.325) -| (2,-1.5) |- (regdst);
    \end{tikzpicture}
    \caption{The Control Unit.}
    \label{fig:cont-unit}
\end{figure}


\subsubsection*{Implementation}
As with the ALU Control, the architecture book~\cite{ref:ark} describes the
logic needed to implement this unit. However, again it is not trivial to extend
later on, so I construct my own Control Unit, and let the VHDL transpiler
handle the logic generation.

I construct a SME process, which has all of the busses as described
previously. The input opcode bus and the ALUOp code bus should both contain an
\texttt{byte} value. The rest of the busses should all contain a \texttt{bool}
value. As with the ALU Control, I have an \texttt{enum} on the opcode, to make
it more human readable. I do not need one for the ALUOp, as the ALU Control
already describes it.

On each clock tick, the Control Unit reads the opcode from the input bus. Then
it should \texttt{switch} on the read opcode, and set all of the flags
accordingly. How the flags should be set, depends on the instruction, and
should be straightforward, but time demanding, as each accepted instruction
needs to be in the \texttt{switch}.

\subsection{Write back}
The final stage of the processor is the Write Back. Here, the values are sent
to the Register File for storing.

\subsubsection*{Implementation}
Usually in the single cycle MIPS processor, there is nothing special in the
Write Back stage. However, in SME it is not allowed to have unclocked cycles,
and there is a cycle from the Register File, through the ALU and the Memory
Unit, and back to the Register File.

To solve this, I introduce a Write Buffer. The write buffer should be clocked,
and takes the Write Data, Write Register and \texttt{WriteEnabled} as input,
and produces the same output. On each clock tick, it should output its stored
values, and store its input values.

I could also have made one of the previous components clocked, however I do
not want to do this, as this specific problem will be solved when I pipeline
the processor in Section~\ref{sec:pipelining}, as the Write Buffer can be
removed.

The Write Buffer and its connections can be seen in Figure~\ref{fig:wb}.
\begin{figure}
    \centering
    \begin{tikzpicture}
        \node[block] (wb) at (0,0) {Write Buffer};
        \node[empty] (wr) at (0,2) {Write Register};
        \node[empty] (wd) at (3,0.5) {Write Data};
        \node[empty] (rw) at (3,-0.5) {\texttt{RegWrite}};
        \node[empty] (wr2) at (-3, 0.5) {Write Register};
        \node[empty] (wd2) at (-3, 0) {Write Data};
        \node[empty] (rw2) at (-3, -0.5) {\texttt{RegWrite}};

        \path[draw, ->] (wr) -- (wb);
        \path[draw, ->] (wd) -| (1.35, 0.45) |- (wb.20);
        \path[draw, color=blue, ->] (rw) -| (1.35, -0.45) |- (wb.340);
        \path[draw, ->] (wb) -- (wd2);
        \path[draw, ->] (wb.160) -| (-1.35,0.45) |- (wr2);
        \path[draw, color=blue, ->] (wb.200) -| (-1.35,-0.45) |- (rw2);
    \end{tikzpicture}
    \caption{The Write Buffer.}
    \label{fig:wb}
\end{figure}
