In this section, I will be looking at some basic combinatorial circuits. I
use this as an entry point for hardware design, both due to the ARK
course~\cite{ref:ark} at DIKU, and due to the architecture
book~\cite{ref:ark-book} recommending to read appendix C, which covers the
basics of logic design, prior to reading chapter 4, which covers implementing a
simple MIPS processor. I am only going to cover a subset of appendix C. It
should be sufficient as an introduction to connecting simpler components
together into a more complex network. It should also be sufficient as an
introduction on going from CSP~\cite{ref:csp} to SME~\cite{ref:sme1,
ref:sme2, ref:sme-cs}.

I start by looking at some logic gates, which implement some basic boolean
functions. Then I will combine these basic gates into more complex networks:
\begin{itemize}
    \item A decoder, which expands an $n$-bit input into $2^n$ outputs.
    \item A half adder, which takes two binary inputs and computes the sum and
        the carry of the two.
    \item A full adder, which does the same operation as the half adder, but
        with an additional third binary input.
    \item A $n$-bit adder, by combining a chain of a single half adder followed
        by $n-1$ full adders.
\end{itemize}

For each of these combinatorial circuits, I go through the theory behind it and
describe the procedure of translating the theory into an SME. All of the
networks follow the same procedure for testing: I construct an SME process,
which sends data on the input busses, and verifies that the data on the output
busses is as expected. Since these networks are very small, I can, in most
cases, test for every possible input. The reason for using this testing
technique is that I can specify the behavior of the processes before
constructing them, and as such don't care how they compute their result,
focussing on the correctness of the result.

\subsection{Basic logic gates}\label{sec:basic}
A logic gate is a circuit abstraction, which has inputs and outputs and
computes the logic function that corresponds to the gates name, i.e. its output
values are based upon the input values. I am going to implement the following
logic gates:

\begin{description}
    \item[\texttt{AND}] - outputs \texttt{1} iff all of its inputs are
        \texttt{1}, otherwise it outputs \texttt{0}.

    \item[\texttt{OR}\ \ ] - outputs \texttt{1} iff one or more of its inputs are
        \texttt{1}, otherwise it outputs \texttt{0}.

    \item[\texttt{NOT}] - outputs the inverse of its input, i.e. \texttt{1}
        becomes \texttt{0} and \texttt{0} becomes \texttt{1}.

    \item[\texttt{XOR}] - outputs \texttt{1} iff exactly one of its inputs are
        \texttt{1}, otherwise it outputs \texttt{0}
\end{description}
The full truth table for all of the four logic gates with 2-bit input
(\texttt{NOT} only looks at Bit1) can be seen in Table~\ref{tab:truth-table}.

\begin{table}
    \centering
    \begin{tabular}{cc|cccc}
        \hline
        \texttt{Bit1} & \texttt{Bit2} & \texttt{AND} & \texttt{OR} &
        \texttt{NOT} & \texttt{XOR} \\
        \hline
        \texttt{0} & \texttt{0} & \texttt{0} & \texttt{0} & \texttt{1} &
            \texttt{0} \\
        \texttt{0} & \texttt{1} & \texttt{0} & \texttt{1} & \texttt{1} &
            \texttt{1} \\
        \texttt{1} & \texttt{0} & \texttt{0} & \texttt{1} & \texttt{0} &
            \texttt{1} \\
        \texttt{1} & \texttt{1} & \texttt{1} & \texttt{1} & \texttt{0} &
            \texttt{0} \\
        \hline
    \end{tabular}
    \caption{The truth table for the four basic logic gates. Note: \texttt{NOT}
    is only considering \texttt{Bit1}.}
    \label{tab:truth-table}
\end{table}

\subsubsection*{Implementation
\footnote{The source code for the Logic Gates is available at~\cite{ref:github}
at \texttt{sme/src/Examples/LogicGates/}}
}
Implementing each of these four logic gates is straightforward: There is an
input bus with two 1-bit values, a process for each of the gates and an output
bus with a 1-bit value for each of the logic gates. I do not need additional
input busses, as each process which uses the bus as input receives its own
copy of the input in each clock, due to the broadcasting in SME~\cite{ref:sme1}.
Furthermore, I only need one output bus, as each process
writes to its own bit within it. Each process takes the two bits from the input
bus, computes its respective logical function and sends the result out onto
its bit on the output bus. The code for the \texttt{AND} gate can be seen in
Listing~\ref{lis:andproc}.

\lstset{language=[Sharp]C}
\noindent
\begin{figure}
\begin{lstlisting}[caption=The SME process for the \texttt{AND} gate., captionpos=b, label=lis:andproc]
public class AND : SimpleProcess
{
    [InputBus]
    Input input;

    [OutputBus]
    Output output;

    protected override void OnTick()
    {
        output.And = input.bit1 && input.bit2;
    }
}
\end{lstlisting}
\end{figure}

I could have made a more complex process (i.e. one that takes the two inputs,
and computes each function). This would reduce the amount of bus connections,
but I want to make each process as simple as possible when constructing the
processor. This is closer to the CSP model, and makes arguing for correctness
simpler. How the logic gate processes and the tester process are connected can
be seen in Figure~\ref{fig:logic-test}. How to run the simulation and generate
the VHDL code can be seen in Listing~\ref{lis:main}.

\begin{figure}
\begin{lstlisting}[caption={The \texttt{Main} function for running an SME simulation, generating the CSV trace file and generating the VHDL code.}, label=lis:main, captionpos=b]
public static void Main(string[] args)
{
    new Simulation()
        .BuildCSVFile()
        .BuildVHDL()
        .Run(typeof(MainClass).Assembly);
}
\end{lstlisting}
\end{figure}

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=1.5cm]
        \node[block] (and) {\texttt{AND}};
        \node[block, right of=and] (or) {\texttt{OR}};
        \node[block, right of=or] (not) {\texttt{NOT}};
        \node[block, right of=not] (xor) {\texttt{XOR}};
        \node[above of=and] (input) at ($(or)!0.5!(not)$) {};
        \node[block, above of=input] (tester) {Tester};

        \path[-] (tester) edge node[midway, right] {input} (input.center);
        \path[draw, ->] (input.center) -| (and.north);
        \path[draw, ->] (input.center) -| (or.north);
        \path[draw, ->] (input.center) -| (not.north);
        \path[draw, ->] (input.center) -| (xor.north);

        \node[below of=and] (output) at ($(or)!0.5!(not)$) {};
        \node[right of=xor] (a) {output};

        \path[draw, -] (and.south) |- (output.center);
        \path[draw, -] (or.south) |- (output.center);
        \path[draw, -] (not.south) |- (output.center);
        \path[draw, -] (xor.south) |- (output.center);

        \path[draw, -] (output.center) -| (a.west);
        \path[draw, ->] (a.west) |- (tester.east);
    \end{tikzpicture}
    \caption{The structure of the test of the Logic Gates.}
    \label{fig:logic-test}
\end{figure}

\subsection{Decoder}
A decoder is a component, which takes an $n$-bit input, and produces an
$2^n$-bit output, where the bit corresponding to the input numbers binary
representation is set to \texttt{1}. E.g. if the input value is the binary
representation of the number 5, then the 5th output bit will be \texttt{1}, and
the rest will be \texttt{0}. Note: the output bits are zero-based, i.e.
there is also an output bit for the binary representation of 0.

A decoder can be made from a set of \texttt{NOT} and \texttt{AND}
gates~\cite{ref:logic}. I need to have $n$ \texttt{NOT} gates and $2^n$
\texttt{AND} gates. I duplicate each input and connect one copy to
a \texttt{NOT} gate. Then for each output I attach an \texttt{AND} gate and
give it inputs corresponding to the binary representation of the number.
E.g. for the number 5, binary representation 101, the 5th \texttt{AND}
gate gets input from Bit0, \texttt{NOT} Bit1 and Bit2. The network of a 2-bit
decoder can be seen in Figure~\ref{fig:2-bit-decoder}.

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=1.5cm]
        \node[block] (and0) {\texttt{AND}};
        \node[block, below of=and0] (and1) {\texttt{AND}};
        \node[block, below of=and1] (and2) {\texttt{AND}};
        \node[block, below of=and2] (and3) {\texttt{AND}};

        \node[right of=and0] (output0) {output0};
        \node[right of=and1] (output1) {output1};
        \node[right of=and2] (output2) {output2};
        \node[right of=and3] (output3) {output3};

        \path[draw, ->] (and0) -- (output0);
        \path[draw, ->] (and1) -- (output1);
        \path[draw, ->] (and2) -- (output2);
        \path[draw, ->] (and3) -- (output3);

        \node[empty, left of=and0] (andinp0) {};
        \node[empty, left of=and1] (andinp1) {};
        \node[empty, left of=and2] (andinp2) {};
        \node[empty, left of=and3] (andinp3) {};

        \node[block, left of=andinp0] (not0) {\texttt{NOT}};
        \node[block, left of=andinp3] (not1) {\texttt{NOT}};

        \node[left of=not0] (input0) {input0};
        \node[left of=not1] (input1) {input1};

        \path[draw, ->] (input0) -- (not0);
        \path[draw, ->] (input1) -- (not1);

        \path[draw, thick, -] (not0) -| (andinp0.155);
        \path[draw, thick, ->] (andinp0.155) -- (and0.155);
        \path[draw, thick, -] (not1.east) -| (andinp0.south);
        \path[draw, thick, ->] (andinp0.south) |- (and0.200);

        \path[draw, ->] (input0) |- (and1.155);
        \path[draw, thick, -] (not1.east) -| (andinp1.340);
        \path[draw, thick, ->] (andinp1.340) -- (and1.200);

        \path[draw, thick, -] (not0.east) -| (andinp2.155);
        \path[draw, thick, ->] (andinp2.155) -- (and2.155);
        \path[draw, ->] (input1.north) |- (and2.200);

        \path[draw, -] (input0) |- (andinp1.295);
        \path[draw, ->] (andinp1.295) |- (and3.155);
        \path[draw, -] (input1) |- (andinp2.200);
        \path[draw, ->] (andinp2.200) |- (and3.200);

    \end{tikzpicture}
    \caption{A 2-bit decoder made by \texttt{AND} and \texttt{NOT} gates.}
    \label{fig:2-bit-decoder}
\end{figure}

\subsubsection*{Implementation
\footnote{The source code for the decoder and the scalable decoder is available
at~\cite{ref:github} in \texttt{sme/src/Examples/Decoder/} and
\texttt{sme/src/Examples/ScalDecoder/}}
}
To implement a 2-bit decoder, I just need to connect logic gate processes,
which I have already constructed in Section~\ref{sec:basic}. How to connect a
2-bit decoder can be seen in Figure~\ref{fig:2-bit-decoder}. I could have made
the decoder a single process. However, I am trying to emphasize connecting
multiple processes together, in order to gain functionality and to be closer to
the CSP model.

However, making a scalable decoder is not trivial, as SME requires everything
to be known at compile time, and I cannot make a generic process, as these
depend on the names of the different busses. To solve this problem, I can use
C\# templates to generate SME code. I chose C\# templates, as I was already
using the C\# version of SME, and as such a C\# template was easy to add. Any
code generator would have been equally good. For each input bit, I create an
input bus. Then, for each input bus, I create an \texttt{NOT} gate process,
which takes the input bus corresponding to its index. Then, I create $2^n$
output busses, and for each of these, I connect an \texttt{AND} gate with its
corresponding output bus. Finally, for each of these \texttt{AND} gates, I
connect the busses whose logical \texttt{AND} will produce a \texttt{1}. E.g.
for \texttt{output0}, I connect all the busses from all of the \texttt{NOT}
gates. For \texttt{output1}, I connect all the \texttt{NOT} gates, except from
the bus from the first \texttt{NOT} gate, as this should be the first input bus.

In order to test the $n$-bit decoder, I also needed to construct the tester
process by using C\# templates, as the tester process also needed a variable
amount of busses.

\subsection{Adder}
An adder is a component, which adds two binary numbers together. As with the
decoder, an adder can be constructed by a combination of \texttt{AND},
\texttt{OR} and \texttt{XOR} gates~\cite{ref:logic}. An $n$-bit adder is a
chain of two major components: a half adder and $n-1$ full
adders~\cite{ref:ark-book}.

The half adder is the initial component in the chain. It takes two binary
inputs, and outputs the sum and the carry of the addition (See
Figure~\ref{fig:half-adder}). The rest of the $n$-bit adder consists of a chain
of $n-1$ full adders that take three inputs A, B, and the carry from the
previous link in the chain, and outputs the sum and the carry of the addition
(See Figure~\ref{fig:full-adder}). The combination of the components can be
seen in Figure~\ref{fig:n-bit-adder}.

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=1.5cm]
        \node[block] (xor) {\texttt{XOR}};
        \node[block, below of=xor] (and) {\texttt{AND}};

        \node[empty, left of=xor] (xorin) {};
        \node[empty, left of=and] (andin) {};

        \node[left of=xorin] (inputa) {InputA};
        \node[left of=andin] (inputb) {InputB};

        \node[empty, right of=xor] (sum) {Sum};
        \node[empty, right of=and] (carry) {Carry};

        \path[draw, -] (inputa) -| (xorin.155);
        \path[draw, ->] (xorin.155) -- (xor.155);
        \path[draw, ->] (xorin.west) |- (and.155);

        \path[draw, thick, -] (inputb.east) -| (xorin.335);
        \path[draw, thick, ->] (xorin.335) -- (xor.205);
        \path[draw, thick, ->] (andin.east) |- (and.205);

        \path[draw, ->] (xor) -- (sum);
        \path[draw, ->] (and) -- (carry);
    \end{tikzpicture}
    \caption{An half adder composed of \texttt{XOR} and \texttt{AND} gates.}
    \label{fig:half-adder}
\end{figure}

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=1.5cm]
        \node[block] (or) {\texttt{OR}};
        \node[empty, left of=or] (orin) {};
        \node[block, left of=orin] (and1) {\texttt{AND}};
        \node[block, above of=and1] (and0) {\texttt{AND}};
        \node[empty, left of=and0] (and0in) {};
        \node[block, above of=and0] (xor1) {\texttt{XOR}};
        \node[empty, left of=xor1] (xor1in) {};
        \node[block, left of=xor1in] (xor0) {\texttt{XOR}};
        \node[empty, left of=xor0] (xor0in) {};
        \node[empty, left of=xor0in] (inputa) {InputA};
        \node[empty, below of=inputa] (inputb) {InputB};
        \node[empty, below of=inputb] (inputc) {InputC};
        \node[empty, right of=inputc] (inputcout) {};
        \node[empty, right of=xor1] (sum) {};
        \node[empty, right of=sum] (summ) {};
        \node[empty, right of=summ] (summm) {Sum};
        \node[empty, right of=or] (carry) {Carry};

        \path[draw, -] (inputa.east) -| (xor0in.155);
        \path[draw, ->] (xor0in.155) -- (xor0.155);
        \path[draw, -] (inputb.east) -| (xor0in.335);
        \path[draw, ->] (xor0in.335) -- (xor0.205);

        \path[draw, -] (inputa.east) -| (inputcout.205);
        \path[draw, ->] (inputcout.205) -- (and1.205);
        \path[draw, -] (inputb.east) -| (inputcout.25);
        \path[draw, ->] (inputcout.25) -- (and1.155);

        \path[draw, thick, -] (inputc.east) -| (and0in.335);
        \path[draw, thick, ->] (and0in.335) -- (and0.205);
        \path[draw, thick, ->] (and0in.335) |- (xor1.205);

        \path[draw, -] (xor0.east) -- (xor1in.west);
        \path[draw, ->] (xor1in.west) |- (xor1.155);
        \path[draw, ->] (xor1in.west) |- (and0.155);

        \path[draw, -] (and1.east) -| (orin.205);
        \path[draw, ->] (orin.205) -- (or.205);
        \path[draw, ->] (xor1) -- (summm);
        \path[draw, -] (and0.east) -| (orin.25);
        \path[draw, ->] (orin.25) -- (or.155);
        \path[draw, ->] (or) -- (carry);
    \end{tikzpicture}
    \caption{A full adder composed of \texttt{AND}, \texttt{OR} and
    \texttt{XOR} gates.}
    \label{fig:full-adder}
\end{figure}

\begin{figure}
    \centering
    \begin{tikzpicture}[node distance=1.5cm]
        \node[block] (half) {Half adder};
        \node[block, below of=half] (full1) {Full adder 1};
        \node[empty, below of=full1] (dot) {...};
        \node[block, below of=dot] (fulln) {Full adder $n-1$};

        \node[empty, left of=half] (spacingl) {};
        \node[empty, right of=half] (spacingr) {};

        \node[empty, left of=spacingl] (in0) {Input 0};
        \node[empty, below of=in0] (in1) {Input 1};
        \node[empty, below of=in1] (vertspacel) {};
        \node[empty, below of=vertspacel] (inn) {Input $n$};

        \node[empty, right of=spacingr] (out0) {Sum 0};
        \node[empty, below of=out0] (out1) {Sum 1};
        \node[empty, below of=out1] (vertspacer) {};
        \node[empty, below of=vertspacer] (outn) {Sum $n-1$};
        \coordinate[below of=fulln] (carry);

        \path[draw, ->] (in0) -- (half);
        \path[draw, ->] (in1) -- (full1);
        \path[draw, ->] (inn) -- (fulln);

        \path[draw, ->] (half) -- (out0);
        \path[draw, ->] (full1) -- (out1);
        \path[draw, ->] (fulln) -- (outn);

        \path[draw, ->] (half) -- node [midway, right] {Carry 0} (full1);
        \path[draw, ->] (full1) -- node [midway, right] {Carry 1} (dot);
        \path[draw, ->] (dot) -- node [midway, right] {Carry $n-2$} (fulln);
        \path[draw, ->] (fulln) -- node [midway, right] {Carry $n-1$} (carry);
    \end{tikzpicture}
    \caption{An $n$-bit adder composed of a half adder, and $n-1$ full adders.
    Note: Input A and B are both inside the inputs for simplicity.}
    \label{fig:n-bit-adder}
\end{figure}

\subsubsection*{Implementation
\footnote{The source code for the half adder, full adder and $n$-bit adder is
available at~\cite{ref:github} in \texttt{sme/src/Examples/Adder/},
\texttt{sme/src/Examples/FullAdder/} and \texttt{sme/src/Examples/Adder32/}}
}
The half adder and the full adder are made like the decoder, in which I have
the basic logic gate processes and connect them as specified in
Figure~\ref{fig:half-adder} and Figure~\ref{fig:full-adder}.

To make the $n$-bit adder, I have to use C\# templates like I did when
constructing the $n$-bit decoder. I program it so that each of the input bits
has its own bus, each of the output sums has its own bus, and each of the
carries has its own bus. I start by making a half adder, which has the
inputs: input A bit 0 and input B bit 0. The initial half adder outputs its sum
on sum 0, and outputs the carry on carry 0. Then I construct $n-1$ full
adders, where full adder $i$ (1-based) has the inputs: input A bit $i$, input
B bit $i$ and carry $i-1$. Each full adder has sum $i$ and carry $i$ as
output.
