In this section, I will be combining the core components into a single cycle
MIPS processor, i.e. a processor where exactly one instruction is executed per
clock cycle. When it is in place, I will be writing the first program, and
compiling it into the processor, and running it.

Following the single cycle MIPS processor, I will be extending
the processor so that it can handle more instructions. Along each added
instruction, I will be extending the first program, in order to verify that
the added instruction works.

Finally, I will be writing two larger programs, and look into compiling them
into a series of hex values, that I can copy straight into the Instruction
Memory.

\subsection{Wiring up the processor}
\begin{figure}
    \centering
    \scalebox{0.5}{
        \begin{tikzpicture}
            \node[block] (reg) at (0,0) {Register file};
            \node[control] (cont) at (-1,3.5) {Control unit};
            \node[block] (jump) at (5,5.5) {Jump unit};
            \node[empty] (splitspace) at (-3,0) {};
            \node[block] (split) at (-4,0) {Splitter};
            \node[block] (if) at (-7,0) {Instruction Memory};
            \node[block] (sign) at (-1,-3) {Sign extend};
            \node[block] (alu) at (4,0) {ALU};
            \node[block] (alucont) at (4,-4) {ALU control};
            \node[block] (mem) at (8.2,0) {Memory};
            \node[mux] (memread) at (10,0) {|};
            \node[mux] (imm) at (3, -0.25) {|};
            \node[mux] (regdst) at (-2,-1) {|};
            \node[block] (pc) at (-8, 2) {PC};
            \node[block] (writebuf) at (6, -2) {Write buffer};

            \path[draw, ->] (if) -- (split);
            \path[draw, -] (split) -- (splitspace.center);
            \path[draw, ->] (splitspace.center) |- (sign);
            \path[draw, ->] (splitspace.center) |- (cont);
            \path[draw, ->] (splitspace.center) |- (reg.160);
            \path[draw, ->] (splitspace.center) |- (reg.200);
            %\path[draw, ->] (splitspace.center) |- (alucont.200);
            \path[draw, ->] (splitspace.center) |- (2,-5.5) -| (alucont.290);
            \path[draw, ->] (splitspace.center) |- (regdst.215);
            \path[draw, ->] (reg.200) -| (-2.5, -0.5) |- (regdst.145);
            %\path[draw, ->] (alucont) -| (12.5, 0) |- (jump);
            \path[draw, thick, ->] (reg.340) -| (2,-0.25) |- (imm.145);
            \path[draw, thick, ->] (2,-0.25) |- (3,0.75) -- (7,0.75) |-
            (mem.164);
            \path[draw, ->] (reg.20) -- (alu.145);
            \path[draw, ->, dashed, color=blue] (alu.20) -| (jump);
            %\path[draw, ->] (alu.340) -- (jal);
            %\path[draw, ->] (jal) -- (mem.196);
            %\path[draw, ->] (jal) -- (writebuf);
            \path[draw, ->] (alu.340) -- (mem.195);
            \path[draw, ->] (imm) -- (alu.202);
            \path[draw, ->] (7, -0.22) |- (8, -1) -| (9.25,-0.5) |-
            (memread.215);
            \path[draw, ->] (mem.10) -- (memread.145);
            \path[draw, ->] (sign) -| (2.5, -1) |- (imm.215);
            \path[draw, ->] (2.5,-3) -| (12, 0) |- (jump.340);
            %\path[draw, thick, ->] (regdst) -| (5, -0.6) |- (jal.210);
            \path[draw, thick, ->] (regdst) -| (writebuf);
            \path[draw, ->] (pc) -| (if);
            \path[draw, ->] (pc) -| (-4, 4) |- (jump);
            \path[draw, ->] (jump.160) -| (pc);
            \path[draw, ->] (memread) -| (11, -1) |- (writebuf.15);
            \path[draw, dashed, ->] (writebuf.170) -| (reg.300);
            \path[draw, dashed, ->] (writebuf) -| (reg);
            \path[draw, dashed, ->, color=blue] (writebuf.190) -| (reg.240);

            \path[draw, ->, color=blue] (alucont) -- (alu);
            \path[draw, ->, color=blue] (cont.35) -| (jump.245);
            \path[draw, ->, color=blue] (cont.25) -| (11.5,0) |-
            (writebuf.345);
            \path[draw, ->, color=blue] (cont.15) -| (memread);
            \path[draw, ->, color=blue] (cont.5) -| (mem.55);
            \path[draw, ->, color=blue] (cont.355) -| (mem.125);
            \path[draw, ->, color=blue] (cont.335) -| (imm);
            \path[draw, ->, color=blue] (cont.325) -| (1.5, -4) |-
            (2, -5) -| (alucont.250);
            \path[draw, ->, color=blue] (cont.240) -| (regdst);

            \node[block] (clock) at (-7, -5.25) {Clock};
            \path[draw, ->, thick, color=green] (clock) -| (-9.5,0) |- (pc);
            \path[draw, ->, thick, color=green] (clock) -| (writebuf);
        \end{tikzpicture}
    }
    \caption{Simple single cycle MIPS processor. The units with | indicate
    multiplexors. The Clock is not an SME process, but has been added to
    emphasize which processes that are clocked. The blue wires indicate control
    wires, i.e. buses that contain a \texttt{bool} value. The black wires
    indicate wires containing data (i.e. \texttt{byte}, \texttt{short} and
    \texttt{uint}). The dashed and bold wires do not indicate anything
    special, but are there to emphasize crossing wires.}
    \label{fig:simple-full}
\end{figure}
The source code for the single cycle processor is available
at~\cite{ref:github} in \texttt{sme/src/Examples/SingleCycleMIPS/}.
With all the components in place, wiring up the single cycle MIPS processor is
straightforward. I just need to declare the busses with the corresponding
names, and then SME handles the wiring process. Had this been implemented in
CSP, each writer process would have to ensure that each reader process would
get a copy. However, as mentioned in the introduction of the paper, SME has
broadcasting channels, which does exactly that so there is no need for
additional logic. Note that as previously mentioned, the Write Buffer and the
PC register has to be clocked processes. The wiring of the single cycle MIPS
processor can be seen in Figure~\ref{fig:simple-full}. Note that the
multiplexors have not been mentioned as components, as these should be simple
and straightforward to insert into the places specified in
Figure~\ref{fig:simple-full}.

\subsection{Writing the first program}
The source code for the first program, with the additional
instructions introduced in Section~\ref{sec:extending}, is available
at~\cite{ref:github} in \texttt{MIPS/first\_program.asm}.
As mentioned before, the first single cycle MIPS processor should be able to
handle \texttt{add}, \texttt{sub}, \texttt{and}, \texttt{or}, \texttt{slt},
\texttt{sw}, \texttt{lw} and \texttt{beq}. As such, the first program should
consist of these. The program and the different parts of each of the
instructions can be seen in Table~\ref{tab:first-program}.
\begin{table}
    \centering
    \begin{adjustbox}{center}
    \begin{tabular}{rllllllll}
        \hline
        Address & Instruction & opcode & rs & rt & rd/imm & shmt & funct & hex \\
        \hline
        \texttt{0x00} & \texttt{add \$3 \$1 \$2} & \texttt{0x00} &
        \texttt{0x01} & \texttt{0x02} & \texttt{0x03} & \texttt{0x00} &
        \texttt{0x20} & \texttt{0x00221820} \\ % 5 + 2 = 7

        \texttt{0x04} & \texttt{sub \$4 \$3 \$2} & \texttt{0x00} &
        \texttt{0x03} & \texttt{0x02} & \texttt{0x04} & \texttt{0x00} &
        \texttt{0x22} & \texttt{0x00622022} \\ %  7 - 2 = 5

        \texttt{0x08} & \texttt{and \$5 \$3 \$1} & \texttt{0x00} &
        \texttt{0x03} & \texttt{0x03} & \texttt{0x05} & \texttt{0x00} &
        \texttt{0x24} & \texttt{0x00612824} \\ % 7 and 5 = 5

        \texttt{0x0C} & \texttt{or \$6 \$3 \$1} & \texttt{0x00} &
        \texttt{0x03} & \texttt{0x03} & \texttt{0x06} & \texttt{0x00} &
        \texttt{0x25} & \texttt{0x00613025} \\ % 7 or 5 = 7

        \texttt{0x10} & \texttt{slt \$7 \$6 \$5} & \texttt{0x00} &
        \texttt{0x06} & \texttt{0x05} & \texttt{0x07} & \texttt{0x00} &
        \texttt{0x2A} & \texttt{0x00C5382A} \\ % 7 < 5 = 1

        \texttt{0x14} & \texttt{sw \$6  0x0(\$0)} & \texttt{0x2B} &
        \texttt{0x00} & \texttt{0x06} & \texttt{0x0000} & - &
        - & \texttt{0xAC060000} \\ % M[0] = 7

        \texttt{0x18} & \texttt{lw \$8  0x0(\$0)} & \texttt{0x23} &
        \texttt{0x00} & \texttt{0x07} & \texttt{0x0000} & - &
        - & \texttt{0x8C070000} \\ % $8 = M[0] (= 7)

        \texttt{0x1C} & \texttt{beq \$5 \$4 0x4} & \texttt{0x04} &
        \texttt{0x05} & \texttt{0x04} & \texttt{0x0004} & - &
        - & \texttt{0x10A40004} \\ % if 5 == 5 then skip next

        \texttt{0x20} & \texttt{add \$9 \$8 \$6} & \texttt{0x00} &
        \texttt{0x08} & \texttt{0x06} & \texttt{0x09} & \texttt{0x00} &
        \texttt{0x20} & \texttt{0x01064820} \\ % 7 + 7 = 14 // should not run!

        \texttt{0x24} & \texttt{add \$10 \$8 \$6} & \texttt{0x00} &
        \texttt{0x08} & \texttt{0x06} & \texttt{0x0A} & \texttt{0x00} &
        \texttt{0x20} & \texttt{0x01065020} \\ % 7 + 7 = 14
        \hline
    \end{tabular}
    \end{adjustbox}
    \caption{The first MIPS program testing the \texttt{add}, \texttt{sub},
    \texttt{and}, \texttt{or}, \texttt{slt}, \texttt{sw}, \texttt{lw} and
    \texttt{beq} instructions. Note that the instruction at \texttt{0x20}
    should be skipped due to the \texttt{beq} at \texttt{0x1C}.}
    \label{tab:first-program}
\end{table}

Inserting this program into the Instruction Memory is straightforward: just
create a \texttt{byte} array, which is initialized to the hex values from the
instruction column in Table~\ref{tab:first-program}. Since I do not have any
way of feeding values into the Register File at this moment, I am going to
hardcode registers 1 and 2 with the values 5 and 2 respectively. When the
program has finished, the contents of the register file should be as in
Table~\ref{tab:first-result}.
\begin{table}
    \centering
    \begin{tabular}{rrrrrrrrrrrrrrrrrrrr}
        \hline
        Address & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \\
        \hline
        Value & 0 & 5 & 2 & 7 & 5 & 5 & 7 & 1 & 7 & 0 & 14 \\
        \hline
    \end{tabular}
    \caption{The register file after the first program has finished.}
    \label{tab:first-result}
\end{table}

\subsection{Extending the accepted instruction set}\label{sec:extending}
Then I will extend the accepted instruction set of the processor, in order to
execute more complex programs. To test each of the added instructions, I just
append the instructions to the initial program.

\subsubsection*{Additional simple R format instructions}
The simplest instructions to add, are the remaining simple R format
instructions. These are \texttt{addu}, \texttt{subu}, \texttt{xor},
\texttt{nor} and \texttt{sltu}. The only modifications are to extend the ALU
Operation \texttt{enum}, the \texttt{funct} \texttt{enum} and then add the
remaining cases in the ALU Control and the ALU. As described in
Section~\ref{sec:alu}, it is important to cast the input \texttt{uint} to
\texttt{int} before making the computation, if the operation is signed.

\subsubsection*{Immediate instructions}
The next instruction I am going to add is the \texttt{ori} instruction, as this
would allow me to feed values into the processor without hardcoding it. While
I am doing this, I also add \texttt{addi}, \texttt{addiu}, \texttt{slti},
\texttt{sltiu}, \texttt{andi} and \texttt{xori}, as these require the same
amount of work.

For the logical immediates, it is important, that the Sign Extend does not
sign extend. As such, I add another signal to the Control Unit:
\texttt{LogicalImmediate}, which goes to the Sign Extend. The Sign Extend
should then, in the case the \texttt{LogicalImmediate} flag is \texttt{1}, cast
its input as \texttt{unsigned}, such that any potential sign bits are not
extended. Then I just need to extend the opcode and ALUOp \texttt{enum}, and
add the remaining \texttt{case}s in the Control Unit and the ALU Control.

Once the processor has been extended, I can prepend instructions at the
beginning of the program, to feed values into the processor. Furthermore, I
can remove the previous hardcoding of initial values in the Register File.

\subsubsection*{Jump instruction}
I am going to introduce another format: the J format~\cite{ref:ark-book}. This
format is used for executing the \texttt{j} instruction. An J format
instruction has two parts:
\begin{itemize}
    \item \makebox[1.6cm]{\texttt{opcode}\hfill}
        \makebox[1.3cm]{(31-26)\hfill} Opcode of the instruction
    \item \makebox[1.6cm]{\texttt{address}\hfill}
        \makebox[1.3cm]{(25-0)\hfill} Jump address
\end{itemize}

The first step is to extend the Splitter, as it should now send the lower 26
bits (25-0) to the Jump Unit. Then the Control Unit should be extended, both in
the opcode \texttt{enum}, and it should have another control signal output:
\texttt{jump}, which should be connected to the Jump Unit.

Finally, the Jump Unit should be extended. It should take the 26 bits from the
Splitter, and left shift it by 2, such that it becomes a 28 bit number. This is
done to pack more information into the address in the instruction, as there are
only 26 bits available. The adresses into the Instruction Memory has to be word
aligned. Since I am making a 32-bit processor, the word size is 4 bytes. As
such, the 2 least significant bits are always 0 to keep the word alignment.

In order to get the address to be a full 32 bit number, the Jump Unit should
take the 4 most significant bits from the PC+4, and prepend them to
the extended number. Finally, I am going to add a multiplexor, which takes
this newly computed address, the previous output address, and the \texttt{jump}
control signal. If the control signal is \texttt{0}, then the original output
should be output, otherwise the newly computed jump address.

\subsubsection*{Branching instruction \texttt{bne}}
Then I am going to add the branch instruction \texttt{bne}. I add another
signal from the Control Unit to the Jump Unit. Then I split the Zero
signal into two, where one of the signals goes to a newly added
\texttt{NOT} gate.  Then I add a multiplexor, which has the \texttt{bne} signal
from the Control Unit as the control signal, and the Zero and the \texttt{NOT}
Zero as inputs. If the \texttt{bne} control signal is \texttt{0}, then the
original Zero should be put on the output, otherwise the \texttt{NOT} Zero. The
output from the multiplexor should go into the \texttt{AND} gate, where the
Zero signal originally went.

\subsubsection*{Additional jump instructions}
Then I am going to add the jump instructions: \texttt{jr} and \texttt{jal},
as these are useful when writing the larger programs later.

I start with \texttt{jr}. The instruction is in R format, so I do not know it
is \texttt{jr}, until it has reached the ALU Control. As such, I am going to
need a control signal from the ALU Control to the Jump Unit. I am also going
to forward Output A from the Register File to the Jump Unit, as this is the
address that the processor should jump to in the \texttt{jr} instruction. The
Jump Unit should not compute the new address in the same manner as with the
\texttt{j} instruction. This is due to the registers being a full 32 bit, and
thus can contain the whole address space. The Jump Unit should also have a
multiplexor, controlling whether the jump address should be the immediate
value, or if it should be the value from the instruction.

For the \texttt{jal} instruction, I am going to need an extra unit following
the ALU. In the case of a \texttt{jal} instruction, the PC+4 address is stored
in register \texttt{31} (which is called the \texttt{\$ra} register).  I add an
additional control signal from the Control Unit: the \texttt{jal} signal. The
new JAL Unit should take three inputs: the ALU Result, the PC+4 and the
\texttt{jal} control signal. It should produce two outputs: the Write Address
for the Write Buffer, and the value to store. If the \texttt{jal} signal is
\texttt{1}, the JAL Unit should output the PC+4 on the value bus, and
\texttt{31} on the address bus. Otherwise it should output the regular ALU
Result, and the regular Write Address.

\subsubsection*{Shift instructions}
Then I am going to add the shift instructions: \texttt{sll}, \texttt{slr},
\texttt{sra}, \texttt{sllv}, \texttt{srlv} and \texttt{srav}, as shifting is
often useful.

The shifting itself is performed in the ALU, and modifying the ALU to handle
these is straightforward. The problem is that in an R format instruction, which
the shift operations are, the shift amount (\texttt{shamt}) is stored in its
own field within the instruction. As such, the splitter should extract these 5
bits, and send them to a new multiplexor, which also takes Output A from the
Register File. As with the \texttt{jr} instruction, I do not know it is a
shifting instruction until it reaches the ALU Control. So to control the new
multiplexor, I need a Control signal from the ALU Control, indicating whether
the multiplexor should output either the \texttt{shamt} or Output A from the
Register File.

\subsubsection*{Multiplication and Division instructions}
Then I am going to add the multiplication and division instructions:
\texttt{mult}, \texttt{multu}, \texttt{div} and \texttt{divu}. All of these
instructions put their result in two special registers: HI and LO. As such, to
get the results from them, I also need to add the \texttt{mfhi}, \texttt{mthi},
\texttt{mflo} and \texttt{mtlo} instructions.

I start by adding the special registers. Since they are performed in the ALU, I
implement them as variables inside the ALU. Then, when the ALU is doing the
computation, it puts the values there, and since the instructions do not write
to registers, touch memory or change the PC register, it does not matter what
is put on the ALU Result or Zero busses.

The instructions handling the moving to and from the HI and LO registers are
fairly simple to implement: just either input or output the corresponding
register, to or from the ALU. The layout for the fully extended single cycle
MIPS processor can be seen in Figure~\ref{fig:single-proc-full}.

\begin{figure}
    \centering
    \scalebox{0.5}{
        \begin{tikzpicture}
            \node[block] (reg) at (0,0) {Register file};
            \node[control] (cont) at (-1,4.5) {Control unit};
            \node[block] (jump) at (5,6.5) {Jump unit};
            \node[empty] (splitspace) at (-3,0) {};
            \node[block] (split) at (-4,0) {Splitter};
            \node[block] (if) at (-7,0) {Instruction Memory};
            \node[block] (sign) at (-1,-3) {Sign extend};
            \node[block] (alu) at (4,0) {ALU};
            \node[block] (alucont) at (4,-4) {ALU control};
            \node[block] (mem) at (8.2,0) {Memory};
            \node[block] (jal) at (6,-0.22) {JAL};
            \node[mux] (memread) at (10,0) {|};
            \node[mux] (shmt) at (2,1.25) {|};
            \node[mux] (imm) at (3, -0.25) {|};
            \node[mux] (regdst) at (-2,-1) {|};
            \node[block] (pc) at (-8, 2) {PC};
            \node[block] (writebuf) at (6, -2) {Write buffer};

            \path[draw, ->] (if) -- (split);
            \path[draw, -] (split) -- (splitspace.center);
            \path[draw, ->] (splitspace.center) |- (sign);
            \path[draw, ->] (splitspace.center) |- (cont);
            \path[draw, ->] (splitspace.center) |- (reg.160);
            \path[draw, ->] (splitspace.center) |- (reg.200);
            \path[draw, ->] (splitspace.center) |- (jump.200);
            \path[draw, ->] (splitspace.center) |- (2,-5.5) -| (alucont.290);
            \path[draw, ->] (splitspace.center) |- (regdst.215);
            \path[draw, ->] (splitspace.center) |- (shmt.145);
            \path[draw, ->] (reg.200) -| (-2.5, -0.5) |- (regdst.145);
            \path[draw, ->, color=blue] (alucont) -| (12.5, 0) |- (jump);
            \path[draw, ->, color=blue] (alucont) -| (shmt);
            \path[draw, thick, ->] (reg.340) -| (2.25,-0.25) |- (imm.145);
            \path[draw, thick, ->] (2.25,-0.25) |- (3,0.75) -- (7,0.75) |-
            (mem.164);
            \path[draw, thick, ->] (reg.20) -| (1.25,0.5) |- (shmt.215);
            \path[draw, thick, -] (1.25,0.5) |- (4,2);
            \path[draw, thick, ->] (4,2) -| (jump.295);
            \path[draw, ->] (shmt) -| (2.5, 0.5) |- (alu.158);
            \path[draw, ->, dashed, color=blue] (alu.20) -| (jump);
            \path[draw, ->] (alu.340) -- (jal);
            \path[draw, ->] (jal) -- (mem.196);
            \path[draw, ->] (imm) -- (alu.202);
            \path[draw, ->] (7, -0.22) |- (8, -1) -| (9.25,-0.5) |-
            (memread.215);
            \path[draw, ->] (mem.10) -- (memread.145);
            \path[draw, ->] (sign) -| (2.5, -1) |- (imm.215);
            \path[draw, ->] (2.5,-3) -| (12, 0) |- (jump.340);
            \path[draw, thick, ->] (regdst) -| (5, -0.6) |- (jal.210);
            \path[draw, ->] (pc) -| (if);
            \path[draw, ->] (pc) -| (-4, 4) |- (jump);
            \path[draw, ->] (jump.160) -| (pc);
            \path[draw, ->] (jal) -- (writebuf);
            \path[draw, ->] (memread) -| (11, -1) |- (writebuf.15);
            \path[draw, dashed, ->] (writebuf.170) -| (reg.300);
            \path[draw, dashed, ->] (writebuf) -| (reg);
            \path[draw, dashed, ->, color=blue] (writebuf.190) -| (reg.240);

            \path[draw, ->, color=blue] (alucont) -- (alu);
            \path[draw, ->, color=blue] (cont.55) -| (jump.220);
            \path[draw, ->, color=blue] (cont.45) -| (jump.230);
            \path[draw, ->, color=blue] (cont.35) -| (jump.245);
            \path[draw, ->, color=blue] (cont.25) -| (11.5,0) |-
            (writebuf.345);
            \path[draw, ->, color=blue] (cont.15) -| (memread);
            \path[draw, ->, color=blue] (cont.5) -| (mem.55);
            \path[draw, ->, color=blue] (cont.355) -| (mem.125);
            \path[draw, ->, color=blue] (cont.345) -| (jal);
            \path[draw, ->, color=blue] (cont.335) -| (imm);
            %\path[draw, ->, color=blue] (cont.325) -| (shmt);
            \path[draw, ->, color=blue] (cont.325) -| (1.5, -4) |-
            (2, -5) -| (alucont.250);
            \path[draw, ->, color=blue] (cont.250) -- (sign.128);
            \path[draw, ->, color=blue] (cont.240) -| (regdst);

            \node[block] (clock) at (-7, -5.25) {Clock};
            \path[draw, ->, thick, color=green] (clock) -| (-9.5,0) |- (pc);
            \path[draw, ->, thick, color=green] (clock) -| (writebuf);
        \end{tikzpicture}
    }
    \caption{Full single cycle MIPS processor. The units with | indicate
    multiplexors. Black wires indicate data wires. Blue wires indicate control
    wires. Green wires indicate clock.}
    \label{fig:single-proc-full}
\end{figure}

\subsection{Larger MIPS programs}
To test the full single cycle MIPS processor, I am going to implement two
programs in MIPS assembly: Quicksort and Towers Of Hanoi. I chose these two
examples, as both are simple to implement, and they do not require anything
special from the environment.

I am not going to give the implementation in MIPS assembly, as this would not
be readable. Instead, I give some low level C code, which should be easy
translatable into MIPS assembly, and easy to verify on a real machine. Once I
have made the assembly, I can easily dump it into MARS~\cite{ref:mars}, which
can then produce a ascii hex dump, which I can then paste into the instruction
memory.

\subsubsection*{Quicksort
\footnote{The source code for the MIPS assembly code and the C code is
available at~\cite{ref:github} in \texttt{MIPS/qsort.asm} and
\texttt{MIPS/qsort.c}}
}
Quicksort is a sorting algorithm, for doing comparison based sorting. There
are three parts of the Quicksort program: Loading data into memory, the
partition function and the quicksort function. I am going to use the
partition function from the algorithm book~\cite{ref:alg}, and the quicksort
function, also from the algorithm book, as both are simple to implement.

I could have implemented the Hoare partitioning algorithm. However, it is not
as simple, and performance is not the target of the program, but rather
correctness of the processor. I.e. that the processor implementation produce
the same result as MARS, in the same amount of clock ticks.

\begin{description}
    \item[Loading data into memory] I start by loading data into memory, so
        that the quicksort program has some data to sort. Which numbers are not
        important, rather the amount of numbers, as the quicksort algorithm
        uses it as argument. I construct a function called \texttt{load}, which
        inserts 8 numbers into the given memory address.
        \lstset{language=[ANSI]C}
        \begin{lstlisting}
 void load(int *a) {
    *(a)   = 5;
    *(a+1) = 8;
    *(a+2) = 2;
    *(a+3) = 9;
    *(a+4) = 1;
    *(a+5) = 3;
    *(a+6) = 7;
    *(a+7) = 11;
}\end{lstlisting}

    \item[The partitioning function] Then I implement the \texttt{partition}
        function as described in the algorithm book~\cite{ref:alg}. Note that
        statements from the algorithm book have been expanded to more closely
        resemble assembly.
\newpage
\begin{lstlisting}
int partition(int *a, int p, int r) {
    int x, i, j, tmp1, tmp2, *addr1, *addr2;
    addr1 = a + r;
    x = *(addr1);
    i = p - 1;
    for (j = p; j < r; j++) {
        addr1 = a + j;
        if (*(addr1) <= x) {
            i++;
            addr1 = a + i;
            addr2 = a + j;
            tmp1 = *(addr1);
            tmp2 = *(addr2);
            *(addr1) = tmp2;
            *(addr2) = tmp1;
        }
    }
    addr1 = a + i + 1;
    addr2 = a + r;
    tmp1 = *(addr1);
    tmp2 = *(addr2);
    *(addr1) = tmp2;
    *(addr2) = tmp1;
    return i + 1;
}
\end{lstlisting}

    \item[The quicksort function] Finally, I implement the
        \texttt{quicksort} function as described in the algorithms book.
\begin{lstlisting}
void quicksort(int *a, int p, int r) {
    if (p < r) {
        int q = partition(a, p, r);
        quicksort(a, p, q-1);
        quicksort(a, q+1, r);
    }
}
\end{lstlisting}

    \item[Initial call to the algorithm] To emphasize the order of calling, and
        the arguments, I also construct a \texttt{main} function.
\begin{lstlisting}
int main() {
    int arr[8], i;
    load(arr);
    quicksort(arr, 0, 7);
}
\end{lstlisting}
\end{description}
Do note that the arguments to both the \texttt{quicksort} and
\texttt{partition} functions, are inclusive. To verify the correctness, we
should look at the memory, where the data now should be in sorted order. The
performance of the quicksort program can be seen in Table~\ref{tab:perf-single}.

\begin{table}
    \centering
    \begin{adjustbox}{center}
    \begin{tabular}{rllllllll}
        & & \texttt{MARS} & & & \texttt{SME} \\
        \hline
        & & \# CT & time (ms) & CR (hz) & \# CT & time (ms) & CR (hz) \\
        \hline
        Towers of Hanoi & $n = 5$ & 719 & 585 & $\sim$1229 & 720 & 516 & $\sim$1395 \\
        Quicksort & $n = 8$ & 483 & 582 & $\sim$829 & 484 & 375 & $\sim$1290 \\
        \hline
    \end{tabular}
    \end{adjustbox}
    \caption{Performance of the Quicksort and Towers of Hanoi programs on the
    single cycle processor. CT is clock ticks. CR is clockrate. Compared to
    \texttt{MARS}, the SME simulation uses an additional clock tick. This is
    due to the simulation using an additional clock tick for shutting down the
    simulation. The execution time of MARS is dominated by the startup time,
    which is why there is no deviation. The benchmark was performed on a laptop
    with an Intel Core i5-5300U (2.3 GHz).}
    \label{tab:perf-single}
\end{table}

\subsubsection*{Towers Of Hanoi
\footnote{The source code for the MIPS assembly code and the C code is
available at~\cite{ref:github} in \texttt{MIPS/hanoi.asm} and
\texttt{MIPS/hanoi.c}}
}
Towers Of Hanoi is a puzzle, where one has to move a tower of discs, from one
peg, to another, with one additional auxilary peg, by only moving one disc at
a time.

By searching on the net, I find a pseudo approach to the problem, which uses
recursion~\cite{ref:hanoi}. I am going to represent the three pegs
as an array, which is three times the size of the tower. As such, each peg is
just one third of the array. Furthermore, as with quicksort, I start by
loading data into memory, have a \texttt{tower} function, which performs the
move, and finally a correct way to call the functions.

\begin{description}
    \item[Loading data into memory] I need to initialize the memory. I am
        going to fill the first third of the array, i.e. the first peg, with
        descending numbers. Each number indicates which disc it is. The rest of
        the pegs with 0, indicating no disc.
\begin{lstlisting}
void init(int num, int *from, int *to, int *aux) {
    int i;
    for (i = 0; i < num; i++) {
        *(from+i) = num - i;
        *(to+i) = 0;
        *(aux+i) = 0;
    }
}
\end{lstlisting}

    \item[The tower function] This is the \texttt{tower} function, which
        moves the discs from peg to peg. I use the same approach as the pseudo
        code~\cite{ref:hanoi}. As with quicksort, the statements have been
        expanded to more closely resemble assembly. I make the three arguments
        as stack pointers to each of the pegs. A stack pointer is the address
        to the next free memory position in the stack data
        structure~\cite{ref:ark-book}. To preserve each of the pointers, I have
        pointers to each of them, i.e. pointer pointer.
\newpage
        \begin{lstlisting}
void tower(int num, int **from, int **to, int **aux) {
    int *t, *f;
    if (num == 0) {
        t = *to;
        f = *from;
        f--;
        *t = *f;
        t++;
        *f = 0;
        *to = t;
        *from = f;
    } else {
        tower(num-1, from, aux, to);
        t = *to;
        f = *from;
        f--;
        *t = *f;
        t++;
        *f = 0;
        *to = t;
        *from = f;
        tower(num-1, aux, to, from);
    }
}
\end{lstlisting}

    \item[Calling the algorithm] I construct a \texttt{main} function, to
        emphasize the order of calling and the arguments.
\begin{lstlisting}
int main() {
    int num = 5;
    int *arr = int[num*3];
    init(num, arr, arr+num, arr+(2*num));
    int *f=num, *t=num, *a=2*num;
    tower(num-1, &f, &a, &t);
}
\end{lstlisting}
\end{description}
When the program has finished, the last third of the array should contain the
numbers in descending order. The performance of the hanoi program can be seen
in Table~\ref{tab:perf-single}.
